Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-04-26 ltx/data_paper/DataPaper.tex
made all Xi variable bold in EM paper

As a result of the model reduction via the basis function decomposition and the decomposition of the connectivity kernel the state space model is formed
\begin{align}
 \mathbf x_{t+1}&=q(\mathbf x_t)\boldsymbol\theta+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r) \\
 \mathbf y_t&=\mathbf C \mathbf x_t+\boldsymbol \epsilon_t,
\end{align}
where the nonlinear function, $q(\cdot)$, is defined as
\begin{equation}\label{eq:QmatrixForSigmapoints}
	q(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
\section{Estimation Algorithm}
Here we describe the problem of estimating non-linear state-space neural field model using expectation-maximisation (EM) based algorithm \cite{Dempster1977,Shumway2000}. The resulting estimation process is an iterative two step procedure that converges on the maximum-likelihood estimates of the parameters. In the E-step we estimate the states $\mathbf x_t$, of the systems and in the M-step we estimate the connectivity kernel parameters $\boldsymbol \theta$, disturbance variance $\sigma_d$ and the observation noise variance $\sigma_{\epsilon}$. Once initialized, at each iteration the sequence of smoothed state vectors is used to update the parameters by maximizing an approximate lower bound on the log likelihood function. The resulting parameters are then used when estimating the new
state vector sequence for the next iteration. The stopping criterion is when the parameter estimates converge. or the log-likelihood variation.
\subsection{E-Step}
In the E-step the additive form of the unscented Rauch-Tung-Striebel smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \cite{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, was set to $10^{-3}$. It is typically set arbitrary small to minimise higher-order effects \cite{Haykin2001}. The other scaling parameter, $\kappa$, was set to $3-n_x$ in accordance to \cite{Julier2002a}.

@@

The goal of the M-step is to maximize a lower bound on the joint likelihood function, $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$, where $\mathbf X$ is the set of all states, $\mathbf Y$ is the set of all observations, and $\boldsymbol \Theta$ denotes the set of all unknown parameters, i.e., connectivity kernel parameters $\boldsymbol\theta$, disturbance variance $\sigma_d$ and the observation noise variance $\sigma_{\epsilon}$.

@@

	\boldsymbol\Theta = \left\lbrace \boldsymbol\theta, \sigma_d, \sigma_{\epsilon}\right\rbrace 

@@

Consider first the maximization of $\mathcal{Q}$-function with respect to $\boldsymbol\theta$, we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\boldsymbol\Xi_4^\top+\boldsymbol\Xi_4)-\frac{2\xi}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
variable $\boldsymbol\Xi_4$ is symmetric and positive-definite by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\boldsymbol\Xi_4^{-1}\left(\boldsymbol\Xi_3^\top-\xi\boldsymbol\Xi_5^\top \right).
\end{equation}
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function. In order to maximise $\mathcal{Q}$-function with respect to $\xi$ we can write
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial\xi}=\frac{2}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -\frac{2\xi}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
setting \eqref{eq:Qpartialxi} to zero and solving for $\xi$
\begin{equation}
 \xi=\frac{\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_3^\top}{\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_5^{\top}}
\end{equation}

Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ gives
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
\begin{align}
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \Xi_5 \boldsymbol\theta 
\end{align}
setting above to zero and solving for $\sigma_e^2$ we have
\begin{align}
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top \Xi_4\boldsymbol\theta+2\xi\Xi_5 \boldsymbol\theta\right]  
\end{align}
The second derivative test confirms this is a maximum.

Computing unknown variables, $\boldsymbol\Theta$, in the maximization step requires calculation of covariance and  cross-covariance terms, including both linear, $\Xi_0$, $\Xi_1$ and $\Xi_2$ and non-linear terms, $\Xi_3$, $\Xi_4$ and  $\Xi_5$. Finding linear $\Xi$-variables is straightforward as they can be obtained using smoother outputs, $\hat{\mathbf x}_t $, $\mathbf P_t$ and $\mathbf M_t$ using the following expressions \cite{Shumway2000} (see supporting information for details)
\begin{align}
	\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0}, \\
\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1}, \\
\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
 \end{align}
However calculation of non-linear $\Xi$-variables required for the M-step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the calculations of the $\Xi_3$, $\Xi_4$ and $\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the estimated state, $\hat{\mathbf x}_t$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the estimated state $\hat{\mathbf{x}}_t$ 
\begin{equation}\label{eq:qTaylor}
 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'
\end{equation}
Define
\begin{equation}
	e(x) = q(x) - q_L(x) \nonumber
\end{equation}

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:qTaylor} into equations \ref{eq:Xi3}-\ref{eq:Xi5} an approximate to the non-linear $\Xi$-variables can be calculated which provides an approximate lower bound on the log-likelihood function (see supporting information), 
\begin{align}
\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}
	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambda}\\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\label{eq:lambdai}\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambdatilde}\\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r' \label{eq:lambdatildei}
\end{align}
Given the state estimates, $\mathbf {\hat x}_t$ the state covariance $\mathbf P_t$ and cross-covariance $\mathbf M_t$ , the $\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.

>>>>

As a result of the model reduction via the basis function decomposition and the decomposition of the connectivity kernel the state space model is formed
\begin{align}
 \mathbf x_{t+1}&=q(\mathbf x_t)\boldsymbol\theta+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r) \\
 \mathbf y_t&=\mathbf C \mathbf x_t+\boldsymbol \epsilon_t,
\end{align}
where the nonlinear function, $q(\cdot)$, is defined as
\begin{equation}\label{eq:QmatrixForSigmapoints}
	q(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
\section{Estimation Algorithm}
Here we describe the problem of estimating non-linear state-space neural field model using expectation-maximisation (EM) based algorithm \cite{Dempster1977,Shumway2000}. The resulting estimation process is an iterative two step procedure that converges on the maximum-likelihood estimates of the parameters. In the E-step we estimate the states $\mathbf x_t$, of the systems and in the M-step we estimate the connectivity kernel parameters $\boldsymbol \theta$, synaptic dynamic $\xi$, disturbance variance $\sigma_d$ and the observation noise variance $\sigma_{\epsilon}$. Once initialized, at each iteration the sequence of smoothed state vectors is used to update the parameters by maximizing an approximate lower bound on the log likelihood function. The resulting parameters are then used when estimating the new
state vector sequence for the next iteration. The stopping criterion is when the parameter estimates converge. or the log-likelihood variation.
\subsection{E-Step}
In the E-step the additive form of the unscented Rauch-Tung-Striebel smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \cite{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, was set to $10^{-3}$. It is typically set arbitrary small to minimise higher-order effects \cite{Haykin2001}. The other scaling parameter, $\kappa$, was set to $3-n_x$ in accordance to \cite{Julier2002a}.

@@

The goal of the M-step is to maximize a lower bound on the joint likelihood function, $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$, where $\mathbf X$ is the set of all states, $\mathbf Y$ is the set of all observations, and $\boldsymbol \Theta$ denotes the set of all unknown parameters, i.e., connectivity kernel parameters $\boldsymbol\theta$, synaptic dynamic $\xi$, disturbance variance $\sigma_d$ and the observation noise variance $\sigma_{\epsilon}$.

@@

	\boldsymbol\Theta = \left\lbrace \boldsymbol\theta, \xi, \sigma_d, \sigma_{\epsilon}\right\rbrace 

@@

Consider first the maximisation of $\mathcal{Q}$-function with respect to $\boldsymbol\theta$, we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\boldsymbol\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\boldsymbol\Xi_4^\top+\boldsymbol\Xi_4)-\frac{2\xi}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
variable $\boldsymbol\Xi_4$ is symmetric and positive-definite by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\boldsymbol\Xi_4^{-1}\left(\boldsymbol\Xi_3^\top-\xi\boldsymbol\Xi_5^\top \right).
\end{equation}
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function. In order to maximise $\mathcal{Q}$-function with respect to $\xi$ we can write
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial\xi}=\frac{2}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -\frac{2\xi}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
setting \eqref{eq:Qpartialxi} to zero and solving for $\xi$
\begin{equation}
 \xi=\frac{\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_3^\top}{\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_5^{\top}}
\end{equation}

Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ gives
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
\begin{align}
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\boldsymbol\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \boldsymbol\Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \boldsymbol\Xi_5 \boldsymbol\theta 
\end{align}
setting above to zero and solving for $\sigma_e^2$ we have
\begin{align}
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\boldsymbol\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top\boldsymbol \Xi_4\boldsymbol\theta+2\xi\boldsymbol\Xi_5 \boldsymbol\theta\right]  
\end{align}
The second derivative test confirms this is a maximum.

Computing unknown variables, $\boldsymbol\Theta$, in the maximization step requires calculation of covariance and  cross-covariance terms, including both linear, $\boldsymbol\Xi_0$, $\boldsymbol\Xi_1$ and $\boldsymbol\Xi_2$ and non-linear terms, $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and  $\boldsymbol\Xi_5$. Finding linear $\boldsymbol\Xi$-variables is straightforward as they can be obtained using smoother outputs, $\hat{\mathbf x}_t $, $\mathbf P_t$ and $\mathbf M_t$ using the following expressions \cite{Shumway2000} (see supporting information for details)
\begin{align}
	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0}, \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1}, \\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
 \end{align}
However calculation of non-linear $\boldsymbol\Xi$-variables required for the M-step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the calculations of the $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and $\boldsymbol\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the estimated state, $\hat{\mathbf x}_t$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the estimated state $\hat{\mathbf{x}}_t$ 
\begin{equation}\label{eq:qTaylor}
 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'
\end{equation}
Define
\begin{equation}
	e(x) = q(x) - q_L(x) \nonumber
\end{equation}

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:qTaylor} into equations \ref{eq:Xi3}-\ref{eq:Xi5} an approximate to the non-linear $\boldsymbol\Xi$-variables can be calculated which provides an approximate lower bound on the log-likelihood function (see supporting information), 
\begin{align}
\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t(\mathbf r)\mathbf P_t \tilde{\Lambda}_t(\mathbf r)\right] \\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}
	 \Lambda_t(\mathbf r)&=\begin{bmatrix}\boldsymbol \lambda_t^{(1)}(\mathbf r) & \boldsymbol \lambda_t^{(2)}(\mathbf r)& \dots &\boldsymbol \lambda_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambda}\\
	 \boldsymbol\lambda_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi^\top(\mathbf r')\tilde{\boldsymbol\Sigma}_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'\label{eq:lambdai}\\
	\tilde{\Lambda}_t(\mathbf r)&=\begin{bmatrix}\tilde{\boldsymbol \lambda}_t^{(1)}(\mathbf r) \\ \tilde{\boldsymbol \lambda}_t^{(2)}(\mathbf r) \\ \vdots \\ \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)\end{bmatrix} \label{eq:Lambdatilde}\\
	 \tilde{\boldsymbol \lambda}_t^{(i)}(\mathbf r)&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r' \label{eq:lambdatildei}
\end{align}
Given the state estimates, $\mathbf {\hat x}_t$ the state covariance $\mathbf P_t$ and cross-covariance $\mathbf M_t$ , the $\boldsymbol\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.
